---
title: "BeerKan experiment and the BEST algorithm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BeerKan experiment and the BEST algorithm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
```

```{r packages, message=FALSE}
library(tidysoilinfiltration)
library(dplyr)
library(tibble)
library(ggplot2)
```

## Overview

The **BeerKan** (Beerkan Estimation of Soil Transfer parameters) protocol is
a low-cost field method: a fixed volume of water is poured repeatedly inside a
ring, and the time for each pour to completely infiltrate is recorded. Unlike
standard ring or Minidisk methods, measurement proceeds at **zero pressure
head** (ponded).

The **BEST algorithm** (Lassabatère et al., 2006) fits the Haverkamp et al.
(1994) quasi-exact implicit infiltration model to the resulting I(t) series,
recovering both the **saturated hydraulic conductivity Ks** and the
**sorptivity S**. The Van Genuchten scale parameter **α** is then recovered
analytically.

| Step | Function |
|------|----------|
| Pour events → cumulative I(t) | `beerkan_cumulative()` |
| VG shape params (n, α) | `infiltration_vg_params(..., suction = 0)` |
| Philip quick estimate | `fit_infiltration()` |
| BEST: Ks, S, α | `fit_best()` |

---

## 1. Field data

A BeerKan run pours a fixed 100 mL of water 10 times into a ring (radius
5 cm), recording how long each pour takes to disappear.

```{r raw-data}
run <- tibble(
  pour   = 1:10,
  volume = 100,                                     # mL per pour
  time   = c(12, 18, 25, 30, 33, 34, 35, 35, 36, 35)  # s to infiltrate
)
run
```

The infiltration time increases early on (soil is initially dry) then
stabilises near 35 s as the soil approaches steady state.

---

## 2. Cumulative infiltration

`beerkan_cumulative()` accumulates both time and volume, then converts to
depth using the ring area.

```{r cumulative}
cum <- beerkan_cumulative(run, volume_col = volume, time_col = time, radius = 5)
cum |> select(pour, volume, time, .cumulative_time, .infiltration)
```

```{r plot-cumulative}
ggplot(cum, aes(x = .cumulative_time, y = .infiltration)) +
  geom_point() +
  geom_line() +
  labs(title = "Cumulative BeerKan infiltration",
       x = "Cumulative time (s)", y = "Infiltration depth (cm)")
```

The curve transitions from a rapid early phase to a linear steady-state,
consistent with the Haverkamp model.

---

## 3. VG shape parameters from texture

For BEST we need the Van Genuchten shape parameter n. Since the experiment
runs at zero pressure head, use `suction = 0` to retrieve only the shape
parameters (`.A` will be `NA`, which is expected):

```{r vg-params}
soil_meta <- tibble(texture = "loam", theta_s = 0.43, theta_i = 0.12) |>
  infiltration_vg_params(texture = texture, suction = 0)

soil_meta |> select(texture, theta_s, theta_i, .n, .alpha, .A)
```

---

## 4. Philip quick estimate

Before running BEST, a quick Philip two-term fit gives a preliminary sorptivity
and conductivity proxy.

```{r philip}
philip <- fit_infiltration(cum,
                           infiltration_col = .infiltration,
                           sqrt_time_col    = .sqrt_time)
philip
```

C₁ (`r signif(philip$.C1, 3)` cm/s) is a rough Ks proxy; BEST will refine
this using the full Haverkamp model.

---

## 5. BEST algorithm

`fit_best()` accepts the cumulative I(t) data together with the soil moisture
parameters (theta_s, theta_i) and the VG shape exponent n.

```{r best}
best <- fit_best(
  cum,
  infiltration_col = .infiltration,
  time_col         = .cumulative_time,
  theta_s          = 0.43,
  theta_i          = 0.12,
  n                = soil_meta$.n
)
best
```

The key outputs:

| Parameter | Symbol | Value | Units |
|-----------|--------|-------|-------|
| Saturated hydraulic conductivity | Ks | `r signif(best$.Ks, 3)` | cm/s |
| Sorptivity | S | `r signif(best$.S, 3)` | cm/s^0.5 |
| VG scale parameter | α | `r signif(best$.alpha, 3)` | 1/cm |

---

## 6. Comparing BEST methods

Three fitting strategies are available. `"steady"` (default) uses the last
`steady_n = 4` points to define the steady-state window; `"slope"` uses the
early-time transient to estimate S independently.

```{r best-methods}
methods <- c("steady", "slope", "intercept")

results <- lapply(methods, function(m) {
  fit_best(
    cum,
    infiltration_col = .infiltration,
    time_col         = .cumulative_time,
    theta_s = 0.43, theta_i = 0.12,
    n       = soil_meta$.n,
    method  = m
  ) |>
    mutate(method = m)
}) |>
  bind_rows()

results |> select(method, .Ks, .S, .alpha, .convergence)
```

---

## 7. Multi-site workflow

For field campaigns, group by site and process all runs together.

```{r multi-site}
field_runs <- tibble(
  site   = rep(c("S1", "S2"), each = 10),
  pour   = rep(1:10, 2),
  volume = 100,
  time   = c(
    12, 18, 25, 30, 33, 34, 35, 35, 36, 35,   # S1 — loam
    8,  11, 14, 17, 19, 20, 20, 21, 20, 21    # S2 — sandy loam (faster)
  )
)

multi_cum <- field_runs |>
  group_by(site) |>
  beerkan_cumulative(volume_col = volume, time_col = time, radius = 5)

multi_best <- multi_cum |>
  group_by(site) |>
  fit_best(
    infiltration_col = .infiltration,
    time_col         = .cumulative_time,
    theta_s = c(0.43, 0.40),
    theta_i = c(0.12, 0.08),
    n       = c(1.56, 1.89)
  )

multi_best |> select(site, .Ks, .S, .alpha, .convergence)
```

Site S2 (sandy loam) shows higher Ks, consistent with its coarser texture and
faster infiltration times.

---

## References

Haverkamp, R., Ross, P. J., Smettem, K. R. J., & Parlange, J.-Y. (1994).
Three-dimensional analysis of infiltration from the disc infiltrometer:
2. Physically based infiltration equation. *Water Resources Research*,
30(11), 2931–2935.

Lassabatère, L., Angulo-Jaramillo, R., Soria Ugalde, J. M., Cuenca, R.,
Braud, I., & Haverkamp, R. (2006). Beerkan estimation of soil transfer
parameters through infiltration experiments—BEST. *Soil Science Society of
America Journal*, 70(2), 521–532.
<https://doi.org/10.2136/sssaj2005.0026>
