---
title: "Minidisk tension-disc infiltrometer workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Minidisk tension-disc infiltrometer workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
```

```{r packages, message=FALSE}
library(tidysoilinfiltration)
library(dplyr)
library(tibble)
library(ggplot2)
```

## Overview

The Minidisk Infiltrometer (Decagon Devices) delivers water at a controlled
tension (suction) below saturation. The Zhang (1997) method extracts the
**unsaturated hydraulic conductivity K(h)** from a two-step analysis:

1. Fit the Philip (1957) two-term polynomial to cumulative infiltration to
   recover the conductivity proxy C₁.
2. Divide C₁ by the soil-specific shape parameter A (derived from Van
   Genuchten parameters and the applied tension) to get K(h).

`minidisk_conductivity()` wraps steps 2 in a single call, giving a four-step
pipeline from raw field readings to K(h):

| Step | Function |
|------|----------|
| Raw readings → I(t) | `infiltration_cumulative()` |
| Philip two-term fit → C₁ | `fit_infiltration()` |
| VG lookup + K(h) = C₁ / A | `minidisk_conductivity()` |

The underlying functions `infiltration_vg_params()`,
`parameter_A_zhang()`, and `hydraulic_conductivity_minidisk()` remain
exported for cases that need more control.

---

## 1. Single-site example

A typical Minidisk run records the reservoir volume (mL) at fixed time
intervals. The disc radius is 2.25 cm (the standard instrument).

```{r single-data}
raw <- tibble(
  time   = seq(0, 300, 30),   # seconds
  volume = c(95, 89, 86, 83, 80, 77, 74, 73, 71, 69, 67)  # mL
)
```

The full pipeline from raw readings to K(h):

```{r single-pipeline}
result <- raw |>
  infiltration_cumulative(time = time, volume = volume) |>
  fit_infiltration(.infiltration, .sqrt_time) |>
  minidisk_conductivity(texture = "sandy loam", suction = 2)

result |> select(.C1, .n, .alpha, .A, .K_h)
```

K(h) ≈ `r signif(result$.K_h, 3)` cm/s at 2 cm tension for this sandy loam
sample.

---

## 2. Multi-site workflow

For field campaigns with multiple samples, group by site before
`infiltration_cumulative()`. Because grouping is now preserved through
cumulative calculations, only one `group_by()` is needed for the whole
pipeline.

```{r multi-data}
multi <- tibble(
  site   = rep(c("A", "B"), each = 11),
  time   = rep(seq(0, 300, 30), 2),
  volume = c(
    95, 89, 86, 83, 80, 77, 74, 73, 71, 69, 67,   # site A — sandy loam
    95, 87, 81, 76, 72, 68, 65, 63, 61, 59, 58    # site B — loamy sand
  )
)

# Per-site metadata (texture, suction)
meta <- tibble(
  site    = c("A", "B"),
  texture = c("sandy loam", "loamy sand"),
  suction = c(2, 2)
)
```

```{r multi-pipeline}
multi_result <- multi |>
  group_by(site) |>
  infiltration_cumulative(time = time, volume = volume) |>
  fit_infiltration(.infiltration, .sqrt_time) |>
  left_join(meta, by = "site") |>
  minidisk_conductivity(texture = texture, suction = suction)

multi_result |> select(site, texture, .C1, .A, .K_h)
```

---

## 3. Analytical A with `method = "zhang"`

For non-standard disc radii or suction levels outside the Decagon table,
pass `method = "zhang"` to compute A analytically from Zhang (1997) rather
than looking it up. The `radius` argument is only used in this mode.

```{r analytical-A}
multi |>
  group_by(site) |>
  infiltration_cumulative(time = time, volume = volume) |>
  fit_infiltration(.infiltration, .sqrt_time) |>
  left_join(meta, by = "site") |>
  minidisk_conductivity(texture = texture, suction = suction,
                        method = "zhang", radius = 2.25) |>
  select(site, texture, .A, .K_h)
```

The analytical A values will differ slightly from the tabulated ones — both
approaches are valid; the choice depends on whether your suction and radius
match the Decagon reference conditions.

---

## 4. Visualisation

```{r multi-plot, message=FALSE}
multi |>
  group_by(site) |>
  infiltration_cumulative(time = time, volume = volume) |>
  ggplot(aes(x = time, y = .infiltration, colour = site)) +
  geom_point() +
  geom_line() +
  labs(
    title  = "Cumulative infiltration — Minidisk",
    x      = "Time (s)",
    y      = "Cumulative infiltration (cm)",
    colour = "Site"
  )
```

---

## References

Decagon Devices, Inc. (2005). *Mini Disk Infiltrometer User's Manual*.

Philip, J. R. (1957). The theory of infiltration: 4. Sorptivity and algebraic
infiltration equations. *Soil Science*, 84(3), 257–264.

Zhang, R. (1997). Determination of soil sorptivity and hydraulic conductivity
from the disk infiltrometer. *Soil Science Society of America Journal*,
61(4), 1024–1030. <https://doi.org/10.2136/sssaj1997.03615995006100060008x>
