---
title: "Minidisk tension-disc infiltrometer workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Minidisk tension-disc infiltrometer workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
```

```{r packages, message=FALSE}
library(tidysoilinfiltration)
library(dplyr)
library(tibble)
library(ggplot2)
```

## Overview

The Minidisk Infiltrometer (Decagon Devices) delivers water at a controlled
tension (suction) below saturation. The Zhang (1997) method extracts the
**unsaturated hydraulic conductivity K(h)** from a two-step analysis:

1. Fit the Philip (1957) two-term polynomial to cumulative infiltration to get
   the conductivity proxy C₁.
2. Divide C₁ by the soil-specific shape parameter A (derived from Van
   Genuchten parameters and the applied tension) to recover K(h).

| Step | Function |
|------|----------|
| Raw readings → I(t) | `infiltration_cumulative()` |
| Philip two-term fit  | `fit_infiltration()` |
| VG parameter lookup  | `infiltration_vg_params()` |
| Analytical A (optional) | `parameter_A_zhang()` |
| K(h) = C₁ / A | `hydraulic_conductivity_minidisk()` |

---

## 1. Single-site example

### 1.1 Field readings

A typical Minidisk run records the reservoir volume (mL) at fixed time
intervals. The disc radius is 2.25 cm (the standard instrument).

```{r single-data}
raw <- tibble(
  time   = seq(0, 300, 30),   # seconds
  volume = c(95, 89, 86, 83, 80, 77, 74, 73, 71, 69, 67)  # mL
)
raw
```

### 1.2 Cumulative infiltration

`infiltration_cumulative()` computes I = ΔV / (π r²) and appends `.sqrt_time`
for use in the Philip fit.

```{r single-cumulative}
cum <- infiltration_cumulative(raw, time = time, volume = volume)
cum
```

### 1.3 Philip two-term fit

```{r single-fit}
philip <- fit_infiltration(cum,
                           infiltration_col = .infiltration,
                           sqrt_time_col    = .sqrt_time)
philip
```

C₁ is the quadratic (time-proportional) coefficient (cm/s); C₂ is the
sorptivity term (cm/s^0.5).

### 1.4 Look up Van Genuchten parameters

The Decagon (2005) lookup table covers 12 USDA texture classes at suction
levels 0.5–7 cm. Assume this soil is sandy loam measured at 2 cm tension.

```{r single-vg}
# Attach texture and suction to the Philip fit result, then look up parameters
result <- philip |>
  mutate(texture = "sandy loam", suction = 2) |>
  infiltration_vg_params(texture = texture, suction = suction)

result |> select(texture, suction, .C1, .n, .alpha, .A)
```

### 1.5 Hydraulic conductivity

```{r single-kh}
result <- hydraulic_conductivity_minidisk(result, C1 = .C1, A = .A)
result |> select(texture, suction, .C1, .A, .K_h)
```

K(h) ≈ `r signif(result$.K_h, 3)` cm/s at 2 cm tension for this sandy loam
sample.

---

## 2. Multi-site workflow

For field campaigns with multiple samples, group the raw data by site and
process everything in a single pipeline.

```{r multi-data}
multi <- tibble(
  site   = rep(c("A", "B"), each = 11),
  time   = rep(seq(0, 300, 30), 2),
  volume = c(
    95, 89, 86, 83, 80, 77, 74, 73, 71, 69, 67,   # site A — sandy loam
    95, 87, 81, 76, 72, 68, 65, 63, 61, 59, 58    # site B — loamy sand
  )
)
```

```{r multi-pipeline}
# Site-level metadata
meta <- tibble(
  site    = c("A", "B"),
  texture = c("sandy loam", "loamy sand"),
  suction = c(2, 2)
)

# infiltration_cumulative() returns an ungrouped tibble, so re-group before
# fit_infiltration() so it fits one model per site.
multi_result <- multi |>
  group_by(site) |>
  infiltration_cumulative(time = time, volume = volume) |>
  group_by(site) |>
  fit_infiltration(infiltration_col = .infiltration,
                   sqrt_time_col    = .sqrt_time) |>
  left_join(meta, by = "site") |>
  infiltration_vg_params(texture = texture, suction = suction) |>
  hydraulic_conductivity_minidisk(C1 = .C1, A = .A)

multi_result |> select(site, texture, .C1, .A, .K_h)
```

---

## 3. Analytical A with `parameter_A_zhang()`

For non-standard disc radii or suction levels not in the Decagon table, compute
A analytically using the Zhang (1997) formula.

```{r analytical-A}
multi_result |>
  select(site, .n, .alpha, suction) |>
  parameter_A_zhang(n = .n, alpha = .alpha, suction = suction) |>
  rename(.A_zhang = .A)
```

The analytical A values should be close to the tabulated `.A` retrieved
via `infiltration_vg_params()`.

---

## 4. Visualisation

```{r multi-plot, message=FALSE}
multi |>
  group_by(site) |>
  infiltration_cumulative(time = time, volume = volume) |>
  ggplot(aes(x = time, y = .infiltration, colour = site)) +
  geom_point() +
  geom_line() +
  labs(
    title  = "Cumulative infiltration — Minidisk",
    x      = "Time (s)",
    y      = "Cumulative infiltration (cm)",
    colour = "Site"
  )
```

---

## References

Decagon Devices, Inc. (2005). *Mini Disk Infiltrometer User's Manual*.

Philip, J. R. (1957). The theory of infiltration: 4. Sorptivity and algebraic
infiltration equations. *Soil Science*, 84(3), 257–264.

Zhang, R. (1997). Determination of soil sorptivity and hydraulic conductivity
from the disk infiltrometer. *Soil Science Society of America Journal*,
61(4), 1024–1030. <https://doi.org/10.2136/sssaj1997.03615995006100060008x>
