---
title: "Ponded ring infiltration workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ponded ring infiltration workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
```

```{r packages, message=FALSE}
library(tidysoilinfiltration)
library(dplyr)
library(tibble)
library(ggplot2)
```

## Overview

Standard ponded ring infiltrometers (single- or double-ring) measure how fast
water infiltrates at saturation. Three empirical models are available:

| Model | Function | Ksat estimate? | Notes |
|-------|----------|----------------|-------|
| Philip (1957) two-term | `fit_infiltration()` | C₁ ≈ Ksat at steady state | Linearisable; fast |
| Horton (1940) | `fit_infiltration_horton()` | fc ≈ Ksat | Requires rates; NLS fit |
| Kostiakov (1932) | `fit_infiltration_kostiakov()` | No | Purely empirical power law |

The processing pipeline is:

1. `infiltration_cumulative()` — convert volume readings to I(t)
2. `infiltration_rate()` — compute interval rates (for Horton)
3. Fit one or more models to the resulting series

---

## 1. Field data

A 10 cm radius ring measured over one hour, reading the reservoir volume (mL)
every six minutes.

```{r raw-data}
ring_raw <- tibble(
  time   = seq(0, 3600, 360),   # seconds (0 – 60 min)
  volume = c(500, 445, 407, 377, 350, 326, 302, 279, 256, 234, 211)  # mL
)
ring_raw
```

## 2. Cumulative infiltration and rates

```{r cumulative}
ring <- ring_raw |>
  infiltration_cumulative(time = time, volume = volume, radius = 10) |>
  infiltration_rate(time_col = time, infiltration_col = .infiltration)

ring |> select(time, .infiltration, .rate, .time_mid)
```

The first `.rate` row is `NA` — there is no preceding interval for the initial
observation.

```{r plot-cumulative}
ggplot(ring, aes(x = time, y = .infiltration)) +
  geom_point() +
  geom_line() +
  labs(title = "Cumulative infiltration — ring infiltrometer",
       x = "Time (s)", y = "Cumulative infiltration (cm)")
```

```{r plot-rate}
ring |>
  filter(!is.na(.rate)) |>
  ggplot(aes(x = .time_mid, y = .rate)) +
  geom_point() +
  geom_line() +
  labs(title = "Infiltration rate — ring infiltrometer",
       x = "Time (s)", y = "Rate (cm/s)")
```

---

## 3. Philip two-term fit

```{r philip}
philip <- fit_infiltration(ring,
                           infiltration_col = .infiltration,
                           sqrt_time_col    = .sqrt_time)
philip
```

C₁ (`r signif(philip$.C1, 3)` cm/s) is a proxy for Ksat under ponded
conditions; C₂ (`r signif(philip$.C2, 3)` cm/s^0.5) is the sorptivity proxy.

---

## 4. Horton model

The Horton (1940) model fits an exponential decay to the infiltration rate
series:

$$f(t) = f_c + (f_0 - f_c) \cdot e^{-k t}$$

where fc ≈ Ksat.

```{r horton}
horton <- fit_infiltration_horton(ring,
                                  rate_col = .rate,
                                  time_col = .time_mid)
horton
```

The Horton estimates: fc = `r signif(horton$.fc, 3)` cm/s ≈ Ksat,
f0 = `r signif(horton$.f0, 3)` cm/s (initial rate),
k = `r signif(horton$.k, 4)` s⁻¹ (decay constant).

Visualise the fitted curve:

```{r horton-plot}
# Predicted rate curve over the observed time range
t_pred <- seq(180, 3420, 60)
pred_horton <- tibble(
  time = t_pred,
  rate = horton$.fc + (horton$.f0 - horton$.fc) * exp(-horton$.k * t_pred)
)

ring |>
  filter(!is.na(.rate)) |>
  ggplot(aes(x = .time_mid, y = .rate)) +
  geom_point() +
  geom_line(data = pred_horton, aes(x = time, y = rate),
            colour = "steelblue", linewidth = 0.8) +
  geom_hline(yintercept = horton$.fc, linetype = "dashed", colour = "steelblue") +
  labs(title    = "Horton model fit",
       subtitle = paste0("fc = ", signif(horton$.fc, 3), " cm/s"),
       x = "Time (s)", y = "Rate (cm/s)")
```

---

## 5. Kostiakov model

The Kostiakov (1932) power model fits cumulative infiltration:

$$I(t) = a \cdot t^b \qquad (a > 0,\ 0 < b < 1)$$

Note that the Kostiakov model does not approach a finite steady-state rate, so
it cannot provide a Ksat estimate.

```{r kostiakov}
kostiakov <- fit_infiltration_kostiakov(ring,
                                        infiltration_col = .infiltration,
                                        time_col         = time)
kostiakov
```

---

## 6. Multi-site workflow

Group the raw data by site and fit all three models in a single pipeline using
`group_by()`.

```{r multi-site}
multi_ring <- tibble(
  site   = rep(c("Field_1", "Field_2"), each = 11),
  time   = rep(seq(0, 3600, 360), 2),
  volume = c(
    500, 445, 407, 377, 350, 326, 302, 279, 256, 234, 211,   # loam
    500, 404, 349, 308, 271, 237, 202, 168, 134, 100,  66    # sandy loam
  )
)

# Cumulative + rate in one pass
multi_cum <- multi_ring |>
  group_by(site) |>
  infiltration_cumulative(time = time, volume = volume, radius = 10) |>
  infiltration_rate(time_col = time, infiltration_col = .infiltration)

# Philip fit
multi_philip <- multi_cum |>
  group_by(site) |>
  fit_infiltration(infiltration_col = .infiltration,
                   sqrt_time_col    = .sqrt_time)
multi_philip

# Horton fit
multi_horton <- multi_cum |>
  group_by(site) |>
  fit_infiltration_horton(rate_col = .rate, time_col = .time_mid)
multi_horton |> select(site, .fc, .f0, .k, .convergence)
```

---

## References

Horton, R. E. (1940). An approach toward a physical interpretation of
infiltration capacity. *Soil Science Society of America Proceedings*,
5, 399–417.

Kostiakov, A. N. (1932). On the dynamics of the coefficient of
water-percolation in soils. *Transactions of the 6th Commission of the
International Society of Soil Science*, Part A, 17–21.

Philip, J. R. (1957). The theory of infiltration: 4. Sorptivity and algebraic
infiltration equations. *Soil Science*, 84(3), 257–264.
